Scalar Feedback based Joint Time-Frequency Precoder interpolation for MIMO-OFDM Systems

### 2019-12-12 paper/main.tex
adding plots and small changed in tex

  In MIMO-OFDM systems, knowledge of the precoding matrix enables
  effective resource allocation for high rates and low BER. However,
  feeding back precoders for all subcarriers introduces large
  overheads. Past work has shown that the orthonormal structure of the
  precoder enables effective parameterization and quantization using
  Givens rotations and Householder transformations. In addition, the
  use of temporal and frequency correlations can further reduce the
  feedback requirements.  We present a framework that utilizes
  adaptive feedback with joint time-frequency prediction for precoder
  construction at the transmitter.  Simulations reveal that the
  proposed method achieves lower BERs for various channel profiles
  compared to other approaches.
\end{abstract}

@@

The use of multiple antennas at the transmitter and receiver can
significantly enhance the performance of wireless systems. Moreover,
in these multiple-input multiple-output (MIMO) wireless systems, the
transmitter can allocate resources more efficiently if the channel
state information (CSI) is made available at the transmitter, both to
improve data rates and to reduce BER~\cite{love2008overview}. To this
end, feeding back the CSI from the receiver to the transmitter
efficiently is necessary. In wireless systems that use orthogonal
frequency division multiplexing (OFDM), the feedback needs to be
provided to the transmitter across all subcarriers, and this places a
large feedback demand. However, since the CSI varies gradually in both
time and frequency, effective techniques to reduce feedback overheads
exist, using frequency domain interpolation temporal prediction. In
particular, knowledge of the precoding matrices (that are typically
the orthonormal right singular vectors of the channel matrix) is
essential to achieve link capacity, as well as to parallelize the
channel. In this paper, we focus on the problem of tracking the
precoders of a MIMO-OFDM system across time and frequency for
efficient CSI feedback by tracking a parametrized version of the matrices.

@@

For the first and the second simulation time instances, since the
precoder is not available at the transmitter, we use 2 bits for
initializing each parameter for effective representation of the
precoder. Therefore, the number of bits for each subcarrier =
$10\times 2$ = 20 (we note that $\phi_0 = \phi_4 = 0$ due to the
non-uniqueness of the SVD). Quantization is performed as given in
Section~\ref{section2}. We note that this is not significant, since,
amortized over long durations, this number becomes small. Subsequent
quantization takes place in the time-domain using one bit for each
parameter. Since the bit budget is small, if the channel varies
sufficiently slowly, either the $\theta_l$ or the $\phi_i$ parameters
are fed back alternately, while retaining the unsent parameters as
being to their previous values, as shown in
Fig.~\ref{fig:adpm-fig}. Here, $\Theta$ is the the collection of all
$\theta_l$s and $\Phi$ is the collection of all $\phi_i$s. This brings
down the bit budget to an average of 5 bits(the length of $\Theta$)
and 7 bits (the length of $\Phi$) over alternate feedback instances,
thereby resulting in an average feedback budget of 5 bits.

@@

Fig.~\ref{fig:error_decay}, shows the rate at which error diminishes
with time for the proposed approach and direct quantization on the
Stiefel manifold using codebooks described
in~\cite{Gupt1905:Predictive}. We see that, since the adaptation of
independent scalar parameters causes the error to decay rapidly, the
error is lower and the convergence is faster.
\begin{figure}
\includegraphics[width=0.5\textwidth]{images/stief_veh.pdf}
\caption{\label{fig:error_decay}A comparison of the $4\times 2$ precoder quantization error vs. time instant for
  the proposed approach (labelled ``TODO'') and direct quantization on
the Stiefel manifold for the Vehicular channel profile.}
\end{figure}

\begin{figure}
\includegraphics[width=0.5\textwidth]{images/cost0035}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 1024
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 3.5\times 10^{-6}$, corresponding to a velocity of 36 km/h.}
\label{fig:ber_ped}
\end{figure}

Fig.~\ref{fig:ber_ped} compares the BER obtained using the proposed
quantization method with that obtained using exact precoders. We can
observe that the performance of the proposed method is very close to
the performance with exact precoders. This can be attributed to the
fact that even the 1 bit update to the scalar parameters is able to
approximate the precoder very accurately. Since the channel variation
across subcarriers is gradual, owing to the limited delay spread of
the channel profile, linear interpolation and prediction yields good
performance. Moreover, this method outperforms the Stiefel manifold
based approach due to the fact that the adaptation is much more
accurate for scalar parameters than the tangent space based adaptation
used over the Stiefel manifold.

Fig.~\ref{fig:ber_veh} shows a similar comparison for a similar
situation with a higher speed. The faster variation of the channel
taps with time necessitates the use of a smaller number of
subcarriers. The smaller number of subcarriers implies that the
predictive quantization and interpolation needs to be performed across
a larger frequency range, thereby potentially reducing the
accuracy. Even in this case, we observe that the scalar parameter
based approach outperforms manifold based quantization by about 5 dB.

\begin{figure}
\includegraphics[width=0.5\textwidth]{images/cost01}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 64
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 1.05\times 10^{-5}$, corresponding to a velocity of 108 km/h.}
\label{fig:ber_veh}
\end{figure}

@@

\section{Conclusion}
\label{section4}
We have presented a predictive quantization based precoder
reconstruction of a MIMO-OFDM system that employs scalar quantization
based feedback of precoder parameters. By quantizing and adapting the
Givens rotation and Householder trasnsformation parameters of the
precoder, the receiver is able to send minimal feedback to the
transmitter and yet enable effective precoder adaptation. Exploiting
the temporal and frequency correlations jointly enhances precoder
accuracy significantly. Simulations reveal that the proposed approach
provides faster convergence and smaller errors in the precoder when
compared to approaches that involve direct quantization on manifold
tangent spaces. Future work would involve tuning the adaptation rate
of the scalar parameters and further reductions of quantization
redundancy to reduce the feedback rates.

>>>>

  In MIMO-OFDM systems, knowledge of the precoding matrix enables
  effective resource allocation for high rates and low BER. However,
  feeding back precoders for all subcarriers introduces large
  overheads. Past work has shown that the orthonormal structure of the
  precoder enables effective parameterization and quantization using
  Givens rotations and Householder transformations. Also, the
  use of temporal and frequency correlations can further reduce the
  feedback requirements. We present a framework that utilizes
  adaptive feedback with joint time-frequency prediction for precoder
  construction at the transmitter.  Simulations reveal that the
  proposed method achieves lower BERs for various channel profiles
  compared to other approaches.
\end{abstract}

@@

The use of multiple antennas at the transmitter and receiver can
significantly enhance the performance of wireless systems. Moreover,
in these multiple-input multiple-output (MIMO) wireless systems, the
transmitter can allocate resources more efficiently if the channel
state information (CSI) is made available at the transmitter, both to
improve data rates and to reduce BER~\cite{love2008overview}. To this
end, it is necessary to efficiently feedback the CSI from the receiver 
to the transmitter. In wireless systems that use orthogonal
frequency division multiplexing (OFDM), the feedback needs to be
provided to the transmitter across all subcarriers, and this places a
large feedback demand. However, since the CSI varies gradually in both
time and frequency, effective techniques to reduce feedback overheads
exist, using frequency domain interpolation temporal prediction. In
particular, knowledge of the precoding matrices (that are typically
the orthonormal right singular vectors of the channel matrix) is
essential to achieve link capacity, as well as to parallelize the
channel. In this paper, we focus on the problem of tracking the
precoders of a MIMO-OFDM system across time and frequency for
efficient CSI feedback by tracking a parametrized version of the matrices.

@@

For the first and the second simulation time instances, since the
precoder is not available at the transmitter, we use 2 bits for
initializing each parameter for effective representation of the
precoder. Therefore, the number of bits for each subcarrier =
$10\times 2$ = 20 (we note that $\phi_0 = \phi_4 = 0$ due to the
non-uniqueness of the SVD). Quantization is performed as given in
Section~\ref{section2}. We note that this is not significant, since,
amortized over long durations, this number becomes small. Subsequent
quantization takes place in the time-domain using one bit for each
parameter. Since the bit budget is small, if the channel varies
sufficiently slowly, either the $\theta_l$ or the $\phi_i$ parameters
are fed back alternately, while retaining the unsent parameters as
being to their previous values, as shown in
Fig.~\ref{fig:adpm-fig}. Here, $\Theta$ is the collection of all
$\theta_l$s and $\Phi$ is the collection of all $\phi_i$s. This brings
down the bit budget to an average of 5 bits(the length of $\Theta$)
and 7 bits (the length of $\Phi$) over alternate feedback instances,
thereby resulting in an average feedback budget of 5 bits.

@@

Fig.~\ref{fig:error_decay}, shows the rate at which error diminishes
with time for the proposed approach and direct quantization on the
Stiefel manifold using codebooks described
in~\cite{Gupt1905:Predictive}. We see that, since the adaptation of
independent scalar parameters causes the error to decay rapidly, the
error is lower, and the convergence is faster.
\begin{figure}
\includegraphics[width=0.5\textwidth]{images/qerror.pdf}
\caption{\label{fig:error_decay}A comparison of the $4\times 2$ precoder quantization error vs. time instant for
  the proposed approach (labelled ``TODO'') and direct quantization on
the Stiefel manifold for the Vehicular channel profile.}
\end{figure}

\begin{figure}
\includegraphics[width=0.5\textwidth]{images/ber1024_035}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 1024
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 3.41\times 10^{-6}$, corresponding to a velocity of 37.8 km/h.}
\label{fig:ber_ped}
\end{figure}

Fig.~\ref{fig:ber_ped} compares the BER obtained using the proposed
quantization method with that obtained using exact precoders. We can
observe that the performance of the proposed method is very close to
the performance with exact precoders. This can be attributed to the
fact that even the 1-bit update to the scalar parameters is able to
approximate the precoder very accurately. Since the channel variation
across subcarriers is gradual, owing to the limited delay spread of
the channel profile, linear interpolation and prediction yields good
performance. Moreover, this method outperforms the Stiefel manifold
based approach due to the fact that the adaptation is much more
accurate for scalar parameters than the tangent space based adaptation
used over the Stiefel manifold.

Fig.~\ref{fig:ber_veh} shows a similar comparison for a similar
situation with a higher speed. The faster variation of the channel
taps with time necessitates the use of a smaller number of
subcarriers. The smaller number of subcarriers implies that the
predictive quantization and interpolation needs to be performed across
a larger frequency range, thereby potentially reducing the
accuracy. Even in this case, we observe that the scalar parameter-based 
approach outperforms manifold based quantization by about 5 dB.

\begin{figure}
\includegraphics[width=0.5\textwidth]{images/ber64_001}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 64
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 1.56\times 10^{-5}$, corresponding to a velocity of 172.8 km/h.}
\label{fig:ber_veh}
\end{figure}

@@

\section{Conclusion}
\label{section4}
We have presented a predictive quantization based precoder
reconstruction of a MIMO-OFDM system that employs scalar quantization
based feedback of precoder parameters. By quantizing and adapting the
Givens rotation and Householder transformation parameters of the
precoder, the receiver can send minimal feedback to the
transmitter and yet enable effective precoder adaptation. Exploiting
the temporal and frequency correlations jointly enhances precoder
accuracy significantly. Simulations reveal that the proposed approach
provides faster convergence and smaller errors in the precoder when
compared to approaches that involve direct quantization on manifold
tangent spaces. Future work would involve tuning the adaptation rate
of the scalar parameters and further reductions of quantization
redundancy to reduce the feedback rates.
